---
title: "filtering sequences using DADA2"
author: "Kimberly Ledger"
date: "2025-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

DADA2 processing of trimmed reads (i.e. primer sequences already removed)

load packages
```{r, warning=FALSE}
library(dada2)
```

when running this code here are the only things that need to be customized
- file path 
- filter parameters and truncate lengths  
- merged sequence length filter 

name file path and check correct files are there - UPDATE FILE PATH DEPENDING ON DATASET - this should point to the trimmed folder
```{r}
path <- "/Users/kimberly.ledger/Documents/noaa-afsc-pinksalmon_herring_stomachs/fastq/trimmed/"
head(list.files(path))
```

create lists for the forward and reverse fastq files 
```{r}
# Forward and reverse fastq filenames have format: SAMPLENAME-N_SX_R1.fastq and SAMPLENAME-N_SX_R2.fastq

fnFs <- sort(list.files(path, pattern="_R1.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2.fastq", full.names = TRUE))

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

## inspect read quality 

visualize quality of forward reads 
```{r, warning=FALSE}
plotQualityProfile(fnFs[1:2])
```

visualize quality of reverse reads
```{r, warning=FALSE}
plotQualityProfile(fnRs[1:2])
```


## filter and trim 

assign filenames 
```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```


**edit this code for trimming**
Use standard filtering parameters: maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2.

mifish sequences: truncLen=c(100,100)
```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(100,100),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=FALSE) # On Windows set multithread=FALSE
head(out)
```


## learn the error rates 
this takes a few minutes to run
```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
```


## subset to include filtFs and filtRs where files exist
```{r}
filtFs <- filtFs[file.exists(filtFs)]
filtRs <- filtRs[file.exists(filtRs)]
```

## apply the sample inference to the filtered and trimmed reads
```{r, results= "hide"}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

take a look at the output
```{r}
dadaFs[[1]]
```

## merge paired reads
```{r, warnings = FALSE, results= "hide"}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE, minOverlap = 8, maxMismatch = 2)
```


Inspect the merger data.frame from the first sample
```{r}
head(mergers[[1]])
```


## construct sequence table 
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

Inspect distribution of sequence lengths
```{r}
table(nchar(getSequences(seqtab)))
```

Remove sequences that are much longer or shorter than expected
```{r}
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% c(164:185)] #mifish
```

Remove chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim)/sum(seqtab2)
```

Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab2), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "correct_length","nonchim")
rownames(track) <- sample.names
head(track)
```

export dada2 filtering summary 
```{r}
out_path <- path <- "/Users/kimberly.ledger/Documents/noaa-afsc-pinksalmon_herring_stomachs/outputs/1_sequence_filtering/"

write.csv(track, file = file.path(out_path, "track.csv"))
```

## export for taxonomic identification
we will want to use the "seqtab.nochim" ASV table for taxonomic analyses
```{r}
write.csv(seqtab.nochim, file.path(out_path, "seqtab.csv"))
```

## also export the seqtab as a Fasta
```{r}
sq <- getSequences(seqtab.nochim)
id <- paste0("ASV", 1:length(sq))

library(seqinr)
write.fasta(sequences = as.list(sq), names=id, file.out = file.path(out_path, "myasvs.fasta"))
```

## and export the seqtab table with the numbered ASVs headers that correspond to those output in the FASTA file.
```{r}
# Make map between brief names and full sequences
briefToSeq <- colnames(seqtab.nochim)
names(briefToSeq) <- paste0("ASV", seq(ncol(seqtab.nochim))) # Seq1, Seq2, ...
# Make new sequence table with brief names
st.brief <- seqtab.nochim
colnames(st.brief) <- names(briefToSeq)

# export the seq table with brief names:
write.csv(st.brief, file= file.path(out_path, "ASVtable.csv"))
```


## create table of ASV labels and ASV sequences
```{r}
asv <- colnames(seqtab.nochim)
asv_id <- paste0("ASV", 1:length(asv))
asv_table <- data.frame(asv, asv_id)

write.csv(asv_table, file = file.path(out_path, "asv_id.csv"))
```

