---
title: "blastn taxonomic analysis"
author: "Kimberly Ledger"
date: "2025-11-14"
output: html_document
---

taxonomic id of blastn output. this code is useful for assigning higher classifications (genus, family, etc) for ASVs that have multispecies hits.   

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
```

## load output from blastn 
```{r load-output-from-taxonkit}
taxonomy <-read.delim("../fastq/trimmed/filtered/outputs/blastnresults", header = FALSE, na.strings=c(""," ","NA"))
head(taxonomy)
colnames(taxonomy) <- c("ASV", "subject", "pident", "length", "mismatch", "gapopen",
  "qstart", "qend", "sstart", "send", "evalue", "bitscore", "qseq", "sseq")
```

## filter the data frame to just work with the curated ASVs (post-LULU)
```{r}
asv_table <- read.csv("~/Documents/noaa-afsc-pinksalmon_herring_stomachs/ref/asvtable_lulu98.csv")

taxonomy <- taxonomy %>%
  filter(ASV %in% asv_table$X)
```


## clean up the header and seperate subject 
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(length > 100) %>% # make sure all retained matches are >100 bp
  select(ASV, subject, pident, length) %>%
  separate(subject, into = c("subject_id", "taxonomy"), sep = "###") %>%
  # Split taxonomy string into ranks
  separate(taxonomy, into = c("root","kingdom","phylum","class","order","family","genus","species"),
           sep = ";", fill = "right") %>%
  select(!root) %>% 
  # Separate name and taxID for each rank
  mutate(across(c(kingdom, phylum, class, order, family, genus),
                ~ str_split_fixed(., "_", 2))) %>% 
  mutate(species = str_match(species, "^(.*_.*)_(\\d+)$")[,2:3]) %>% 
  mutate(kingdom = kingdom[,1],
    phylum  = phylum[,1],
    class   = class[,1],
    order   = order[,1],
    family  = family[,1],
    genus   = genus[,1],
    species = species[,1]) %>%
  select(ASV, pident, length, kingdom, phylum, class, order, family, genus, species) %>%
  unique() %>%
  filter(!str_detect(species, "_x_")) %>% ##remove hybrids
  mutate(species_name = str_replace_all(species, "_", " "))   
```

## now look at the fish and figure out what fish taxa are in/out of our range
```{r}
to_check_range <- tax_df %>%
  filter(class == "Actinopteri") %>%
  ungroup() %>%
  select(species, genus, family, order, class, phylum) %>%
  unique() %>% 
  mutate(species_name = str_replace_all(species, "_", " "))
```

## check ranges for species using rfishbase 
```{r}
#library(remotes)
#remotes::install_github("ropensci/rfishbase")
library(rfishbase)

#one-time download of all fishbase tables... this takes a bit 
#fb_import()

#first, validate species names using rfishbase synonyms
spp_df <- synonyms(to_check_range$species_name)

#check the synonyms
syn <- spp_df %>% 
  filter(Status == "synonym")

#rename any synonyms with the validated name  
to_check_range_validated <- to_check_range %>%
  rename(synonym = species_name) %>%
  left_join(syn, by = "synonym") %>%
  mutate(synonym = ifelse(is.na(Species), synonym, Species)) %>%
  select(genus:synonym) %>%
  rename(species = synonym) %>%
  unique()

#get distribution info 
spp_distribution <- ecosystem(to_check_range_validated$species) %>%
  select(Species, EcosystemName) %>%
  unique()

#add column to designate if we will consider a species as "in range"- for this study, this will be NE Pacific 
spp_distribution_range <- spp_distribution %>%
  mutate(in_range = ifelse(is.na(EcosystemName), NA, "no"),
         #in_range = ifelse(EcosystemName == "Pacific Ocean", "yes", in_range),
         in_range = ifelse(EcosystemName == "Gulf of Alaska", "yes", in_range))

#keep just a list of spp names and yes/no/NA for "in range"  - this way we can keep track of what spp didn't have any reference information in fishbase to determine range 
in_range_df <- spp_distribution_range %>%
  filter(in_range == "yes")
```

check to make sure the range filter worked okay
```{r}
removed <- anti_join(to_check_range_validated, in_range_df, by = c("species" = "Species"))
```

for now, i'm being more restrictive and using the GofA filter. might consider including all Pacific Ocean species? 


Polypera greeni - Liparis greeni 
ASV 68 - Sus scrofa 
remove the non-fish... 


## what non-fish are here? 
```{r}
not_Actinopteri <- tax_df %>%
  filter(class != "Actinopteri") 

unique(not_Actinopteri$species)
```

## keep only the "in range" fish species in the taxon dataframe - and all non-fish - and the positive control species (Acipenser fulvescens)
```{r}
in_range_spp <- unique(in_range_df$Species)

taxa_keepers <- c(in_range_spp, unique(not_Actinopteri$species), "Acipenser fulvescens", "Ammodytes personatus")

taxon_df <- tax_df %>%
  filter(species_name %in% taxa_keepers)
```


## figure out what ASVs are unambiguously a single species
```{r}
single_spp_seqs <- taxon_df %>% 
  group_by(ASV) %>%
  add_tally(name = "n_taxa") %>%
  filter(n_taxa == 1)

#modify headers to match with filtering to follow 
single_spp_seqs <- single_spp_seqs %>%
  rename(seq_percID = pident) %>%
  mutate(id_rank = 1,
         top_perc = seq_percID,
         diff = 0) %>%
  select(!length) %>%
  select(!species_name) %>%
  select(!n_taxa)
```

## remove the single-species seqs from the dataframe and then rank the hits by % identity for the remaining seqs
```{r}
seq_id_diff <- taxon_df %>%
  anti_join(., single_spp_seqs) %>%  ## comment out this line if there are no single species ASVs
  select(-length) %>%
  group_by(ASV, species, genus, family, order, class, phylum, kingdom) %>%
    mutate(seq_percID = max(pident)) %>%
    group_by(ASV, species, genus, family, order, class, phylum, kingdom, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(ASV) %>%
      mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
        arrange(diff)

seq_id_diff %>%
  filter(diff > 0)
```

Now you have the single best entry for each species for each sequence ranked and with the difference between the first and second ranked entries calculated.

For sequences with multiple top hits, where the difference between ranked taxa < 0.5, you will end up defaulting to genus- or family-level ID. 

## Figure out which differences are > 0.5% and eliminate those first
```{r}
# filter out any taxa that are >2% less matching identity than the top taxonomic hit for a given sequence
to_remove_low_perc_hits <- seq_id_diff %>%
  ungroup() %>%
  group_by(ASV) %>%
  filter(diff > 0.5)

keepers <- seq_id_diff %>%
  anti_join(to_remove_low_perc_hits)
```

## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(ASV) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  bind_rows(single_spp_seqs) %>% # combine the single spp data
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  group_by(ASV, genus) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single genus
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = genus)
  

## Family-level matches
single_family <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  group_by(ASV, family) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single family
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "family") %>%
  mutate(taxon = family)


## Order-level matches
single_order <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  group_by(ASV, order) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single order
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "order") %>%
  mutate(taxon = order)


## Class-level matches
single_class <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  group_by(ASV, class) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single class
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "class") %>%
  mutate(taxon = class)


## Phylum-level matches
single_phylum <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(ASV, phylum) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single phylum
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "phylum") %>%
  mutate(taxon = phylum)

# Kingdom level matches
single_kingdom <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(ASV, kingdom) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single kingdom
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "kingdom") %>%
  mutate(taxon = kingdom)
```


## recombine the full data set now that the appropriate level of taxonomy has been determined
```{r}
sorted_tax_df <- bind_rows(single_kingdom, single_phylum, single_class, single_order, single_family, single_genus, singleton_df)
```

## Create output taxonomy data frames
```{r}
uncollapsed_taxonomy <- sorted_tax_df %>%
  select(-top_perc, -id_rank) %>%
  unique() %>%
  arrange(ASV)

# and then collapse that down to just a single taxon per ASV
collapsed_taxonomy <- uncollapsed_taxonomy %>%
  select(ASV, taxon, taxonomic_level) %>%
  unique()
```

```{r}
write.csv(uncollapsed_taxonomy, "../outputs/uncollaspsed_taxonomy.csv")
write.csv(collapsed_taxonomy, "../outputs/collaspsed_taxonomy.csv")
```
