---
title: "Decontamination of MiFish Amplicons"
author: "Kimberly Ledger"
output: html_document
date: "2025-11-14"
---

libraries
```{r}
library(tidyverse)
library(stringr)
library(reshape2)
rename <- dplyr::rename
library(here)
```

load the lulu curated asv table 
```{r}
asv_table <- read.csv(here("outputs/2_lulu", "asvtable_lulu98.csv")) 
#562 ASVs across 280 samples

#transpose 
asv_table <- data.frame(t(asv_table))

#set column names to be ASV# 
colnames(asv_table) <- asv_table[1,]

#remove row that has ASV#
asv_table <- asv_table[-1, ]

#make sure reads are numbers
# Convert all character columns to numeric
for (col in names(asv_table)) {
  asv_table[[col]] <- as.numeric(asv_table[[col]])
}

asv_table$sample_ID <- rownames(asv_table)  #make make sample ID a column 

#reorder columns
asv_table <- asv_table %>%
  select(sample_ID, everything()) 

#lulu output changed "_" to "."
asv_table$sample_ID <- gsub("\\.", "-", asv_table$sample_ID) 
```

load taxonomic assignments 
```{r}
taxonomy <- read.csv(here("outputs/3_blastn_taxonomy", "collaspsed_taxonomy.csv"))
```

## decon 1 - remove ASVs that do not have a taxonomic assignment 
```{r}
asv_table_long <- asv_table %>%
  pivot_longer(cols = -1, names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) 

asv_table_filter1 <- asv_table_long %>%
  filter(ASV %in% taxonomy$ASV)

removed_asvs <- anti_join(asv_table_long, asv_table_filter1) %>%
  group_by(ASV) %>%
  summarize(total_reads = sum(reads)) %>%
  arrange(desc(total_reads))
removed_asvs

#how many ASVs and total reads removed because of no tax assignment? 
nrow(removed_asvs)
sum(removed_asvs$total_reads)
#proportion of reads removed because of no taxonomic assignment
sum(removed_asvs$total_reads)/sum(asv_table_long$reads)
```

add column to sample tracking table to include number of reads per sample after taxonomic assignment 
```{r}
track <- read.csv(here("outputs/1_sequence_filtering", "track.csv")) %>%
  rename(sample_ID = X)

reads_per_sample <- asv_table_filter1 %>%
  group_by(sample_ID) %>%
  summarize(tax_reads = sum(reads))

track <- track %>%
  left_join(reads_per_sample)
```

at first glance this is all looking reasonable 

load sample metadata 
```{r}
metadata <- read.csv("../pinksalmon_and_herring_metadata.csv") %>%
  rename(sample_ID = extraction_ID) %>%
  rename(stomach = Species)

## _ were changed to - by illumina output
metadata$sample_ID <- gsub("_", "-", metadata$sample_ID) 
```

add column to the ASV table that labels the sample type and taxonomic assignment
```{r}
asv_table_with_sample_type <- metadata %>%
  dplyr::select(sample_ID, FishID, stomach, sequence_date, sample_type) %>%
  left_join(asv_table_filter1, by = "sample_ID") %>%
  left_join(taxonomy, by = "ASV")
```

## Step 1. Account for tag-jumping by using the positive controls  

first, get a feel for the data by visualizing positive controls 
```{r}
asv_table_with_sample_type %>%
  filter(sample_type == "positive") %>%
  ggplot(aes(x=sample_ID, y=reads, fill=ASV)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  facet_wrap(~sequence_date, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in positive control") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_with_sample_type %>%
  filter(sample_type == "positive") %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  facet_wrap(~sequence_date, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "taxon reads in positive control") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

subtract the proportion of reads that jumped into the positive control samples from each environmental sample 

identify the maximum proportion of reads for each ASV found in the positive controls by sequencing run 
** excluding the high prop of pink salmon in PC27
```{r}
prop_asvs_in_positives <- asv_table_with_sample_type %>%
  filter(sample_type == "positive") %>%
  group_by(sample_ID) %>%
  mutate(TotalReadsPerSample = sum(reads)) %>%
  mutate(Prop = reads/TotalReadsPerSample) %>%
  filter(!(sample_ID == "PC-27" & taxon == "Oncorhynchus_gorbuscha")) %>%
  group_by(sequence_date, ASV) %>%
  summarise(max_prop = max(Prop)) %>%
  arrange(desc(max_prop)) 
prop_asvs_in_positives
```


subtract the max proportion of tag-jumped reads for each ASV from samples
```{r}
indexhop_table <- asv_table_with_sample_type %>%
  group_by(sample_ID) %>%
  mutate(TotalReadsPerSample = sum(reads, na.rm = T)) %>%
  left_join(prop_asvs_in_positives, by = c("ASV", "sequence_date")) %>%
  mutate(IndexHoppingReads = TotalReadsPerSample*max_prop) %>%
  mutate(IndexHoppingReads = if_else(is.na(IndexHoppingReads), 0, IndexHoppingReads)) %>%
  mutate(IndexHoppingReads = as.integer(IndexHoppingReads)) %>%
  mutate(reads_IndexHop_removed = reads - IndexHoppingReads) %>%
  mutate(reads_IndexHop_removed = if_else(reads_IndexHop_removed < 0, 0, reads_IndexHop_removed))
head(indexhop_table)
```

clean up the table by removing columns no longer needed 
```{r}
asv_table_filter2 <- indexhop_table %>%
  dplyr::select(!reads) %>%
  dplyr::select(!TotalReadsPerSample) %>%
  dplyr::select(!max_prop) %>%
  dplyr::select(!IndexHoppingReads) %>%
  dplyr::rename(reads = reads_IndexHop_removed)
```


## Step 2. Account for contaminants in negative controls 

plot negatives 
```{r}
asv_table_filter2 %>%
  filter(reads > 0) %>%
  filter(sample_type != "sample") %>%
  filter(sample_type != "positive") %>%
  ggplot(aes(x=sample_ID, y=reads, fill=ASV)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  facet_wrap(sample_type~sequence_date, scales = "free") + 
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads in negative controls") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )

asv_table_filter2 %>%
  filter(reads > 0) %>%
  filter(sample_type != "sample") %>%
  filter(sample_type != "positive") %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  facet_wrap(sample_type~sequence_date, scales = "free") + 
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "taxa reads in negative controls") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

summarize read depth by sample type
```{r}
asv_table_filter2 %>%
  group_by(sample_ID, sample_type) %>%
  summarize(tot_reads = sum(reads)) %>%
  group_by(sample_type) %>%
  summarize(
    min   = min(tot_reads, na.rm = TRUE),
    q25   = quantile(tot_reads, 0.25, na.rm = TRUE),
    median = median(tot_reads, na.rm = TRUE),
    mean   = mean(tot_reads, na.rm = TRUE),
    q75   = quantile(tot_reads, 0.75, na.rm = TRUE),
    max   = max(tot_reads, na.rm = TRUE)
  )
```


perhaps come back to this to include a filter? 

## Step 4. Remove low read depth samples based on ASV accumulation curve

```{r}
library(vegan)

asv_table_wide <- asv_table_filter2 %>%
  select(sample_ID, ASV, reads) %>%
  mutate(reads = as.integer(reads)) %>%
  pivot_wider(names_from = ASV, values_from = reads, values_fill = 0)

sample_IDs <- asv_table_wide$sample_ID

asv_table_wide <- asv_table_wide %>%
  ungroup() %>%
  select(-sample_ID)

## plots the figure
rarecurve(asv_table_wide, step = 20, col = "blue", label = FALSE, 
          main = "Sequencing Effort Curves",
          xlab = "Sequencing Depth", ylab = "Number of ASVs Identified",
          xlim = c(0,5000))
```

the asv vs seq depth curves plateau at a pretty low read depth (which is good)   

summarize in a table how many pcr replicates meet certain read count thresholds 
```{r}
read_summary <- asv_table_filter2 %>%
  group_by(sample_ID, sample_type) %>%
  summarize(tot_reads = sum(reads)) %>%
  arrange(desc(tot_reads)) %>%
  group_by(sample_type) %>%
  summarize(atleast0 = sum(tot_reads >= 0),
            atleast100 = sum(tot_reads >= 100),
            atleast500 = sum(tot_reads >= 500),
            atleast1k = sum(tot_reads >= 1000),
            atleast2k = sum(tot_reads >= 2000))
read_summary
```

threshold of 1000 reads per pcr replicate seems appropriate 

```{r}
asv_table_filter3 <- asv_table_filter2 %>%
  group_by(sample_ID) %>%
  mutate(tot_reads = sum(reads)) %>%
  filter(tot_reads >= 1000)
```


```{r}
saveRDS(asv_table_filter3, file =  here("outputs/4_decontamination", "decontaminated_asv_table.RDS"))
```


### how many reads does each asv with taxon assignment have? 
```{r}
asv_table_filter3 %>%
  group_by(taxon, ASV) %>%
  summarize(total_reads = sum(reads)) %>%
  arrange(desc(total_reads))

asv_table_filter3 %>%
  filter(sample_type == "sample") %>%
  group_by(taxon, sequence_date) %>%
  summarize(total_reads = sum(reads)) %>%
  pivot_wider(names_from = "sequence_date", values_from = "total_reads") 
  #arrange(desc(total_reads))
```


