---
title: "analysis of stomach contents"
output: html_document
date: "2025-11-14"
---

```{r}
library(tidyverse)
```


load decontaminated ASV table w/ taxon assignments 
```{r}
taxon_table <- readRDS(here("outputs/4_decontamination", "decontaminated_asv_table.RDS")) %>%
  filter(sample_type == "sample") %>%
  mutate(year = ifelse(sequence_date == "2023_11_15", 2023, 2024)) %>%
  filter(class != "Mammalia") %>% ## remove contamination from mammals 
  group_by(sample_ID, FishID, stomach, year, taxon) %>%
  summarize(reads = sum(reads)) %>%
  filter(taxon != "Acipenser_fulvescens") ## remove positive control 
```

compare sequencing depth by year
```{r}
taxon_table %>%
  group_by(year, sample_ID) %>%
  summarize(tot_reads = sum(reads)) %>%
  ungroup()%>%
  group_by(year) %>%
  summarize(
    min   = min(tot_reads, na.rm = TRUE),
    q25   = quantile(tot_reads, 0.25, na.rm = TRUE),
    median = median(tot_reads, na.rm = TRUE),
    mean   = mean(tot_reads, na.rm = TRUE),
    q75   = quantile(tot_reads, 0.75, na.rm = TRUE),
    max   = max(tot_reads, na.rm = TRUE))
```


## Pacific herring stomach contents 
```{r}
herring <- taxon_table %>%
  filter(stomach == "PacificHerring") %>% 
  filter(reads > 0)

total_samples_per_year <- herring %>%
  group_by(year) %>%
  summarise(n_total = n_distinct(sample_ID), .groups = "drop")

herring_detections <- herring %>%
  group_by(year, taxon) %>%       # group by species
  summarise(n_samples = n(), .groups = "drop") %>%
  left_join(total_samples_per_year, by = "year") %>%
  mutate(prop_samples = n_samples/n_total)
```

look closer at the proportion of herring reads from herring stomachs
```{r}
herring_in_herring <- herring %>%
  group_by(sample_ID) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  filter(taxon == "Clupea_pallasii") %>%
  mutate(non_herring_reads = total_reads - reads)

ggplot(herring_in_herring, aes(x=reads)) + 
  geom_histogram(binwidth = 500, fill = "steelblue", color = "black") +
  labs(x = "herring reads in herring",
    y = "Count"
  ) +
  theme_minimal()

ggplot(herring_in_herring, aes(x=prop_reads)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(x = "proportion of herring reads in herring",
    y = "Count"
  ) +
  theme_minimal()

herring_discards_1 <- herring_in_herring %>%
  filter(prop_reads > 0.95) 

herring_discards_2 <- herring_in_herring %>%
  filter(non_herring_reads < 1000)
```

exclude samples if 95% of total reads are herring OR less than 1000 non-herring reads (doesn't have to meet both criteria, just one) 
```{r}
herring_filtered <- herring %>%
  filter(!sample_ID %in% herring_discards_1$sample_ID) %>%
  filter(!sample_ID %in% herring_discards_2$sample_ID)

herring_filtered %>%
  group_by(year) %>%
  summarize(n_extracts = n_distinct(sample_ID),
            n_fish = n_distinct(FishID))
```


out of curiousity, plot the distribution of pink salmon reads in herring stomachs 
```{r}
tmp <- herring_filtered %>%
  group_by(sample_ID) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  filter(taxon != "Clupea_pallasii") %>%
  mutate(total_reads_nonhost = sum(reads),
         prop_reads_nonhost = reads/total_reads_nonhost) %>%
  filter(taxon == "Oncorhynchus_gorbuscha")

ggplot(tmp, aes(x=reads)) + 
  geom_histogram(binwidth = 500, fill = "steelblue", color = "black") +
  labs(x = "Pink salmon reads",
    y = "Count"
  ) +
  theme_minimal()

ggplot(tmp, aes(x=prop_reads)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(x = "Proportion (pink salmon/total reads)",
    y = "Count"
  ) +
  theme_minimal()

ggplot(tmp, aes(x=prop_reads_nonhost)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(x = "Proportion (pink salmon/nonhost reads)",
    y = "Count"
  ) +
  theme_minimal()
```

double check 24-05-25 and 24-05-195 are actually herring stomachs (and not pink salmon)? 

set a threshold that a species needs to have at least 1000 non-host reads (filter included above) and >=1% reads (non-host) to be considered present in the stomach?
```{r}
herring_diet <- herring_filtered %>%
  filter(taxon != "Clupea_pallasii") %>% 
  group_by(sample_ID) %>%
  mutate(total_reads_nonhost = sum(reads),
         prop_reads_nonhost = reads/total_reads_nonhost) %>%
  filter(prop_reads_nonhost >= 0.01)

total_samples_per_year_keepers <- herring_diet %>%
  group_by(year) %>%
  summarise(n_total = n_distinct(sample_ID), .groups = "drop")

herring_diet_taxa <- herring_diet %>%
  group_by(year, taxon) %>%
  summarize(n_samples = n(), .groups = "drop") %>%
  left_join(total_samples_per_year_keepers, by = "year") %>%
  mutate(prop_samples = n_samples/n_total)

herring_diet_taxa_wide <- herring_diet_taxa %>%
  select(!n_total) %>%
  pivot_wider(
    names_from = year,
    values_from = c(n_samples, prop_samples), values_fill = 0) %>%
  arrange(desc(prop_samples_2023))
```

make list of detected prey (>5% in a given year, ie. at least 4 detections in 2023 or at least 2 detections in 2024)
```{r}
herring_diet_common <- herring_diet_taxa %>%
  filter(prop_samples > 0.05) %>%
    select(!n_total) %>%
  pivot_wider(
    names_from = year,
    values_from = c(n_samples, prop_samples), values_fill = 0) %>%
  arrange(desc(prop_samples_2023))
```

mean and range of taxa detected per stomach 
```{r}
herring_diet %>%
  group_by(year, sample_ID) %>%
  summarise(n = n_distinct(taxon)) %>%
  group_by(year) %>%
  summarise(min = min(n),
            max = max(n),
            mean = mean(n))
```

mean proportion of reads of common species (when detected?)
```{r}
#2023 
herring_diet_common_2023 <- herring_diet_common %>%
  filter(n_samples_2023 > 0)

#2024
herring_diet_common_2024 <- herring_diet_common %>%
  filter(n_samples_2024 > 0)

#2023 
reads_2023 <- herring_diet %>%
  filter(year == 2023) %>%
  filter(taxon %in% herring_diet_common_2023$taxon) %>%
  group_by(taxon) %>%
  summarize(
    min   = min(prop_reads_nonhost, na.rm = TRUE),
    q2.5 = quantile(prop_reads_nonhost, 0.025, na.rm = TRUE),
    q25   = quantile(prop_reads_nonhost, 0.25, na.rm = TRUE),
    median = median(prop_reads_nonhost, na.rm = TRUE),
    mean   = mean(prop_reads_nonhost, na.rm = TRUE),
    q75   = quantile(prop_reads_nonhost, 0.75, na.rm = TRUE),
    q97.5 = quantile(prop_reads_nonhost, 0.975, na.rm = TRUE),
    max   = max(prop_reads_nonhost, na.rm = TRUE)
  )
reads_2023

ggplot(reads_2023, aes(y = taxon)) +
  geom_errorbarh(aes(xmin = min, xmax = max), size = 0.5, height = 0, color = "black") +  # min-max line
  geom_errorbarh(aes(xmin = q25, xmax = q75), size = 2, height = 0, color = "black") +
  geom_point(aes(x = median), size = 3, color = "black") + 
  #geom_point(aes(x = mean), size = 3, color = "grey") +                            
  labs(x = "proportion of reads per stomach", y = "Taxon", 
       title = "2023 Pacific herring stomach content") +
  scale_y_discrete(limits = rev) +
  theme_minimal()
  
tmp23 <- reads_2023 %>%
  select(taxon, mean)

herring_diet_common_2023 %>%
  select(taxon, n_samples_2023, prop_samples_2023) %>%
  left_join(tmp23)

#2024 
reads_2024 <- herring_diet %>%
  filter(year == 2024) %>%
  filter(taxon %in% herring_diet_common_2024$taxon) %>%
  group_by(taxon) %>%
  summarize(
    min   = min(prop_reads_nonhost, na.rm = TRUE),
    q2.5 = quantile(prop_reads_nonhost, 0.025, na.rm = TRUE),
    q25   = quantile(prop_reads_nonhost, 0.25, na.rm = TRUE),
    median = median(prop_reads_nonhost, na.rm = TRUE),
    mean   = mean(prop_reads_nonhost, na.rm = TRUE),
    q75   = quantile(prop_reads_nonhost, 0.75, na.rm = TRUE),
    q97.5 = quantile(prop_reads_nonhost, 0.975, na.rm = TRUE),
    max   = max(prop_reads_nonhost, na.rm = TRUE)
  )
reads_2024

tmp24 <- reads_2024 %>%
  select(taxon, mean)

herring_diet_common_2024 %>%
  select(taxon, n_samples_2024, prop_samples_2024) %>%
  left_join(tmp24) %>%
  arrange(desc(n_samples_2024))
```

plot
```{r}
library(colorspace)
colors_42 <- qualitative_hcl(42, palette = "Dark 3")  # HCL-based qualitative palette

herring_plot <- herring_diet %>%
  ggplot(aes(x=sample_ID, y=prop_reads_nonhost, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
  scale_fill_manual(values = colors_42) +
  facet_wrap(~year, scales = "free_x", ncol = 1) + 
   labs(
    y = "proportion of sequencing reads",
    x = "FishID",
    title = "herring stomach content")  +
  scale_x_discrete(labels = setNames(herring_diet$FishID, herring_diet$sample_ID)) +  # replace sample_ID labels with FishID
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  ) 

herring_plot
```

save plot 
```{r}
ggsave(here("outputs/5_stomachs", "herring_plot.png"), plot = herring_plot, width = 12, height = 6, dpi = 300)
```

export herring diet dataframe 
```{r}
saveRDS(herring_diet, file = here("outputs/5_stomachs", "herring_stomachs.RDS"))
```


## Pink salmon stomach contents 
```{r}
pink <- taxon_table %>%
  filter(stomach == "PinkSalmon") %>%
  filter(reads > 0)

total_samples_per_year <- pink %>%
  group_by(year) %>%
  summarise(n_total = n_distinct(sample_ID), .groups = "drop")

pink_detections <- pink %>%
  group_by(year, taxon) %>%       # group by species
  summarise(n_samples = n(), .groups = "drop") %>%
  left_join(total_samples_per_year, by = "year") %>%
  mutate(prop_samples = n_samples/n_total)
```

look closer at the proportion of pink reads from pink stomachs
```{r}
pink_in_pink <- pink %>%
  group_by(sample_ID) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  filter(taxon == "Oncorhynchus_gorbuscha") %>%
  mutate(non_pink_reads = total_reads - reads)

ggplot(pink_in_pink, aes(x=reads)) + 
  geom_histogram(binwidth = 500, fill = "steelblue", color = "black") +
  labs(x = "pink reads in pink",
    y = "Count"
  ) +
  theme_minimal()

ggplot(pink_in_pink, aes(x=prop_reads)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(x = "proportion of pink reads in pink",
    y = "Count"
  ) +
  theme_minimal()

pink_discards_1 <- pink_in_pink %>%
  filter(prop_reads > 0.95) 

pink_discards_2 <-  pink_in_pink %>%
  filter(non_pink_reads < 1000)
```

exclude samples if 95% of total reads are pink OR less than 1000 non-pink reads (doesn't have to meet both criteria, just one) 
```{r}
pink_filtered <- pink %>%
  filter(!sample_ID %in% pink_discards_1$sample_ID) %>%
  filter(!sample_ID %in% pink_discards_2$sample_ID)

pink_filtered %>%
  group_by(year) %>%
  summarize(n_extracts = n_distinct(sample_ID),
            n_fish = n_distinct(FishID))
```


plot the distribution of herring reads in pink salmon stomachs 
```{r}
tmp <- pink_filtered %>%
  group_by(sample_ID) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  filter(taxon != "Oncorhynchus_gorbuscha") %>%
  mutate(total_reads_nonhost = sum(reads),
         prop_reads_nonhost = reads/total_reads_nonhost) %>%
  filter(taxon == "Clupea_pallasii")

ggplot(tmp, aes(x=reads)) + 
  geom_histogram(binwidth = 500, fill = "steelblue", color = "black") +
  labs(x = "Herring reads",
    y = "Count"
  ) +
  theme_minimal()

ggplot(tmp, aes(x=prop_reads)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(x = "Proportion (herring/total reads)",
    y = "Count"
  ) +
  theme_minimal()

ggplot(tmp, aes(x=prop_reads_nonhost)) + 
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black") +
  labs(x = "Proportion (herring/nonhost reads)",
    y = "Count"
  ) +
  theme_minimal()
```

double check 24-06-468, 24-06-284, 24-06-451, and others(?) are actually pink salmon stomachs (and not herring)

maybe set a threshold that a species needs to have >=1% reads (non-host) to be considered present in the stomach?  
```{r}
pink_diet <- pink_filtered %>%
  filter(taxon != "Oncorhynchus_gorbuscha") %>% 
  filter(taxon != "Oncorhynchus_1") %>%
  group_by(sample_ID) %>%
  mutate(total_reads_nonhost = sum(reads),
         prop_reads_nonhost = reads/total_reads_nonhost) %>%
  filter(prop_reads_nonhost > 0.01)

total_samples_per_year_keepers <- pink_diet %>%
  group_by(year) %>%
  summarise(n_total = n_distinct(sample_ID), .groups = "drop")

pink_diet_keepers <- pink_diet %>%
  group_by(year, taxon) %>%
  summarize(n_samples = n(), .groups = "drop") %>%
  left_join(total_samples_per_year_keepers, by = "year") %>%
  mutate(prop_samples = n_samples/n_total)

pink_diet_keepers_wide <- pink_diet_keepers %>%
  select(!n_total) %>%
  pivot_wider(
    names_from = year,
    values_from = c(n_samples, prop_samples), values_fill = 0) %>%
  arrange(desc(prop_samples_2023))
```

make list of detected prey (>5% in a given year, ie. at least 4 detections in 2023 or at least 2 detections in 2024)
```{r}
pink_diet_common <- pink_diet_keepers %>%
  filter(prop_samples > 0.05) %>%
    select(!n_total) %>%
  pivot_wider(
    names_from = year,
    values_from = c(n_samples, prop_samples), values_fill = 0) %>%
  arrange(desc(prop_samples_2023))
```

mean and range of taxa detected per stomach 
```{r}
pink_diet %>%
  group_by(year, sample_ID) %>%
  summarise(n = n_distinct(taxon)) %>%
  group_by(year) %>%
  summarise(min = min(n),
            max = max(n),
            mean = mean(n))
```

mean proportion of reads of common species (when detected?)
```{r}
#2023 
pink_diet_common_2023 <- pink_diet_common %>%
  filter(n_samples_2023 > 0)

#2024
pink_diet_common_2024 <- pink_diet_common %>%
  filter(n_samples_2024 > 0)

#2023 
reads_2023 <- pink_diet %>%
  filter(year == 2023) %>%
  filter(taxon %in% pink_diet_common_2023$taxon) %>%
  group_by(taxon) %>%
  summarize(
    min   = min(prop_reads_nonhost, na.rm = TRUE),
    q2.5 = quantile(prop_reads_nonhost, 0.025, na.rm = TRUE),
    q25   = quantile(prop_reads_nonhost, 0.25, na.rm = TRUE),
    median = median(prop_reads_nonhost, na.rm = TRUE),
    mean   = mean(prop_reads_nonhost, na.rm = TRUE),
    q75   = quantile(prop_reads_nonhost, 0.75, na.rm = TRUE),
    q97.5 = quantile(prop_reads_nonhost, 0.975, na.rm = TRUE),
    max   = max(prop_reads_nonhost, na.rm = TRUE)
  )
reads_2023

ggplot(reads_2023, aes(y = taxon)) +
  geom_errorbarh(aes(xmin = min, xmax = max), size = 0.5, height = 0, color = "black") +  # min-max line
  geom_errorbarh(aes(xmin = q25, xmax = q75), size = 2, height = 0, color = "black") +
  geom_point(aes(x = median), size = 3, color = "black") + 
  #geom_point(aes(x = mean), size = 3, color = "grey") +                            
  labs(x = "proportion of reads per stomach", y = "Taxon", 
       title = "2023 pink salmon stomach content") +
  scale_y_discrete(limits = rev) +
  theme_minimal()
  
tmp23 <- reads_2023 %>%
  select(taxon, mean)

pink_diet_common_2023 %>%
  select(taxon, n_samples_2023, prop_samples_2023) %>%
  left_join(tmp23)

#2024 
reads_2024 <- pink_diet %>%
  filter(year == 2024) %>%
  filter(taxon %in% pink_diet_common_2024$taxon) %>%
  group_by(taxon) %>%
  summarize(
    min   = min(prop_reads_nonhost, na.rm = TRUE),
    q2.5 = quantile(prop_reads_nonhost, 0.025, na.rm = TRUE),
    q25   = quantile(prop_reads_nonhost, 0.25, na.rm = TRUE),
    median = median(prop_reads_nonhost, na.rm = TRUE),
    mean   = mean(prop_reads_nonhost, na.rm = TRUE),
    q75   = quantile(prop_reads_nonhost, 0.75, na.rm = TRUE),
    q97.5 = quantile(prop_reads_nonhost, 0.975, na.rm = TRUE),
    max   = max(prop_reads_nonhost, na.rm = TRUE)
  )
reads_2024

tmp24 <- reads_2024 %>%
  select(taxon, mean)

pink_diet_common_2024 %>%
  select(taxon, n_samples_2024, prop_samples_2024) %>%
  left_join(tmp24) %>%
  arrange(desc(n_samples_2024))
```

plot
```{r}
colors_19 <- qualitative_hcl(19, palette = "Dark 3")  # HCL-based qualitative palette

pink_plot <- pink_diet %>%
  ggplot(aes(x=sample_ID, y=prop_reads_nonhost, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
  scale_fill_manual(values = colors_19) +
  facet_wrap(~year, scales = "free_x", ncol = 1) + 
   labs(
    y = "proportion of sequencing reads",
    x = "FishID",
    title = "pink salmon stomach content")  +
  scale_x_discrete(labels = setNames(pink_diet$FishID, pink_diet$sample_ID)) +  # replace sample_ID labels with FishID
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  ) 

pink_plot


```

save plot 
```{r}
ggsave(here("outputs/5_stomachs", "pink_plot.png"), plot = pink_plot, width = 12, height = 6, dpi = 300)
```

export prey detection table for pink salmon stomachs 
```{r}
saveRDS(pink_diet, file = here("outputs/5_stomachs", "pinksalmon_stomachs.RDS"))
```

look at the metadata
```{r}
metadata <- read.csv("../pinksalmon_and_herring_metadata.csv") %>%
  rename(sample_ID = extraction_ID) %>%
  rename(stomach = Species)

herring_meta <- metadata %>%
  filter(stomach == "PacificHerring")

herring_meta %>%
  group_by(collection_year) %>%
  summarise(sites = unique(location1))

herring_meta %>%
  group_by(collection_year) %>%
  summarize(n_fish = n_distinct(FishID))

herring_meta %>%
  group_by(collection_year) %>%
  summarize(n_fish = n_distinct(sample_ID))
```

```{r}
pink_meta <- metadata %>%
  filter(stomach == "PinkSalmon")

pink_meta %>%
  group_by(collection_year) %>%
  summarise(sites = unique(location1))

pink_meta %>%
  group_by(collection_year) %>%
  summarize(n_fish = n_distinct(FishID))

pink_meta %>%
  group_by(collection_year) %>%
  summarize(n_fish = n_distinct(sample_ID))
```

