---
title: "analysis of stomach contents"
output: html_document
date: "2025-11-14"
---

```{r}
library(tidyverse)
```


load decontaminated ASV table w/ taxon assignments 
```{r}
taxon_table <- readRDS("../outputs/decontaminated_asv_table.RDS") %>%
  filter(sample_type == "sample") %>%
  mutate(year = ifelse(sequence_date == "2023_11_15", "y_2023", "y_2024")) %>%
  group_by(sample_ID, FishID, Species, year, taxon) %>%
  summarize(reads = sum(reads)) %>%
  filter(taxon != "Acipenser_fulvescens") ## remove positive control 
```


## Pacific herring stomach contents 
```{r}
herring <- taxon_table %>%
  filter(Species == "PacificHerring") 

herring_taxa <- herring %>%
  group_by(year, taxon) %>%
  summarise(total = sum(reads)) %>%
  pivot_wider(names_from = "year", values_from = "total")

1639262/sum(herring_taxa$y_2023, na.rm = T) #prop of herring reads in 2023 data
541001/sum(herring_taxa$y_2024, na.rm = T) #prop of herring reads in 2024 data
```

identify the 10 most common taxa in herring stomachs 
```{r}
herring_mostcommon_2023 <- herring %>%
  filter(year == "y_2023") %>%
  group_by(taxon) %>%
  summarise(total = sum(reads)) %>%
  arrange(desc(total)) %>%
  filter(total > 10000)

herring_mostcommon_2024 <- herring %>%
  filter(year == "y_2024") %>%
  group_by(taxon) %>%
  summarise(total = sum(reads)) %>%
  arrange(desc(total)) %>%
  filter(total > 6000)

herring_mostcommon <- herring_mostcommon_2023 %>%
  bind_rows(herring_mostcommon_2024) %>% select(taxon) %>% unique()

herring <- herring %>%
  mutate(taxon_common = ifelse(taxon %in% herring_mostcommon$taxon, taxon, "rare"))
```


plot samples WITH herring reads
```{r}
herring %>%
  filter(reads > 0) %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "sequencing reads",
    x = "sample ID",
    title = "herring stomachs with herring reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  
```

```{r}
herring %>%
  filter(reads > 0) %>%
  filter(taxon_common != "Clupea_pallasii") %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "sequencing reads",
    x = "sample ID",
    title = "herring stomachs WITHOUT herring reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  
```

let's say a sample need >1000 non-herring reads in order to establish its diet 
```{r}
herring_diet <- herring %>%
  filter(taxon != "Clupea_pallasii") %>%
  group_by(sample_ID) %>%
  mutate(ReadsPerSample = sum(reads)) %>%
  filter(ReadsPerSample > 1000) %>%
  mutate(read_prop = reads/ReadsPerSample)
```

```{r}
herring_diet %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "sequencing reads",
    x = "sample ID",
    title = "herring stomachs WITHOUT herring reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  

herring_diet %>%
  ggplot(aes(x=sample_ID, y=read_prop, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "proportion of sequencing reads",
    x = "sample ID",
    title = "herring stomachs WITHOUT herring reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  
```



## pink salmon stomach contents 
```{r}
pink <- taxon_table %>%
  filter(Species == "PinkSalmon") 

pink_taxa <- pink %>%
  group_by(year, taxon) %>%
  summarise(total = sum(reads)) %>%
  pivot_wider(names_from = "year", values_from = "total")

1314325/sum(pink_taxa$y_2023, na.rm = T) #prop of pink reads in 2023 data
920777/sum(pink_taxa$y_2024, na.rm = T) #prop of pink reads in 2024 data
```

identify the 10 most common taxa in pink salmon stomachs 
```{r}
pink_mostcommon_2023 <- pink %>%
  filter(year == "y_2023") %>%
  group_by(taxon) %>%
  summarise(total = sum(reads)) %>%
  arrange(desc(total)) %>%
  filter(total > 200)

pink_mostcommon_2024 <- pink %>%
  filter(year == "y_2024") %>%
  group_by(taxon) %>%
  summarise(total = sum(reads)) %>%
  arrange(desc(total)) %>%
  filter(total > 1000)

pink_mostcommon <- pink_mostcommon_2023 %>%
  bind_rows(pink_mostcommon_2024) %>% select(taxon) %>% unique()

pink <- pink %>%
  mutate(taxon_common = ifelse(taxon %in% pink_mostcommon$taxon, taxon, "rare"))
```


plot samples WITH herring reads
```{r}
pink %>%
  filter(reads > 0) %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "sequencing reads",
    x = "sample ID",
    title = "pink stomachs with pink reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  
```

```{r}
pink %>%
  filter(reads > 0) %>%
  filter(taxon_common != "Oncorhynchus_gorbuscha") %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "sequencing reads",
    x = "sample ID",
    title = "pink stomachs WITHOUT pink reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  
```

let's say a sample need >1000 non-herring reads in order to establish its diet 
```{r}
pink_diet <- pink %>%
  filter(taxon != "Oncorhynchus_gorbuscha") %>%
  group_by(sample_ID) %>%
  mutate(ReadsPerSample = sum(reads)) %>%
  filter(ReadsPerSample > 1000) %>%
  mutate(read_prop = reads/ReadsPerSample)
```

```{r}
pink_diet %>%
  ggplot(aes(x=sample_ID, y=reads, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "sequencing reads",
    x = "sample ID",
    title = "pink stomachs WITHOUT pink reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  

pink_diet %>%
  ggplot(aes(x=sample_ID, y=read_prop, fill=taxon_common)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "proportion of sequencing reads",
    x = "sample ID",
    title = "pink stomachs WITHOUT pink reads")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 8)
  )  
```




